Toward a clearer understanding of the inverse association between cancer and dementia

### 2021-04-23 06_presentations/ser2021/index.Rmd
SER presentation

---
title: "‘Dissecting the causal question underlying the association between cancer and dementia"
subtitle: ""  
author: 
  - "L. Paloma Rojas-Saunero"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["r", "css"]
      countIncrementalSlides: FALSE
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

@@

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#011A5E",
  header_font_google = google_font("Fira Sans", "500"),
  table_row_even_background_color = "#FFFFFF",
  table_border_color = "#011A5E",
  text_font_google   = google_font("Lato", "400", "400i"),
  text_font_size = "25px"
)
```

background-image: url(./figs/erasmus.png)
background-position: 5% 95%
background-size: 25%, 15%
class: center, middle

## Dissecting the causal question underlying the association between cancer and dementia

@@

## Introduction

.pull-left[

Observational studies have found a protective effect of cancer diagnosis in the risk of Alzheimer´s disease or related dementias (ADRD).

These studies lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD
]

@@

.pull-right[ 
![](figs/forest_plot.PNG)
]

.footnote[Ospina-Romero M. JAMA Network Open. 2020]

## Cancer diagnosis as an exposure

```{r, fig.align='center', out.width= "40%"}
include_graphics("figs/causal_risks.PNG")

```
.footnote[L. Frain et al. Alzheimer’s & Dementia. 2017]
---

.pull-left[
![](figs/pin_knockout.PNG) 
]

.pull-right[

<br><br><br>

![](figs/tment_pin1.PNG)

.pull-right[
.footnote[Driver J. et al. Biochimica et Biophysica Acta. 2015.]
]
---

class: middle, center

## How can we conclude on causal pathways with observational data?

@@

- To progressively build a causal directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.

- To exemplify different scenarios with real data collected from the Rotterdam Study.

- To describe the challenges and how they translate into the analytic decisions.

### Causal question

.pull-left[

What is the effect of a drug targeted to Pin1 $(D)$ in the risk of ADRD at time *t* $(Y_{t})$  compared to standard treatments?

We can imagine an RCT where middle-aged participants with signs of cognitive decline are recruited.

@@

@@

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$D$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (p) -- (y);
\end{tikzpicture}
```
]

---
class: middle

.pull-left[
- The drug is not available yet, but we have better chances of having Pin1 expression as a biomarker in a closer future.

- This biomarker could be measured in stored blood samples in a cohort study

- We can set a threshold for _Pin1 over-expression_ = $P$

- Since this is within an observational setting we will have to address confounding $(L)$

@@

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (l) -- (p);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (y);
\end{tikzpicture}
```
]

class: middle

.pull-left[

- Pin1 over-expression stimulates tumor growth

- Tumor growth leads to cancer diagnosis

- Cancer diagnosis $(P^*)$ represents a proxy for $P$

- We are assuming that the captured effect of $P^*$ and $Y_{t}$ is only through $P$
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->, violet] (p) -- (pstar);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);

\end{tikzpicture}
```
]

@@

Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life.

---
class:middle

.pull-left[

- In the cohort setting people will develop cancer over time

- You need to survive long enough to have a cancer diagnosis $(D_k = 0)$

@@

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\end{tikzpicture}
```
]

class:middle

.pull-left[

- We need to isolate the effect of $P$ in $P^*$ and mimic a setting where over time those who died had would have had the same risk of developing cancer over time based on additional information $C_1, C_2$

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) -- (d);
\draw[->] (c) -- (pstar);
\draw[->] (c) -- (d);
\end{tikzpicture}
```
]

---
class: middle, center

## Overview of the causal question and assumptions related to death as a competing event of ADRD

---
class: middle

.pull-left[

- We are interested in the direct effect of cancer in dementia, not mediated through death

- We can consider death as a censoring event that could be prevented

- We rely on the independent censoring assumption

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```
]

---
class: middle

.pull-left[

- But we cannot consider $D_{t}$ and $Y_{t+1}$ as independent events unconditionally

- We need information on time-varying factors that increase the risk of dying and having dementia over time $C$

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```

@@

---
class: middle

.pull-left[

- We also have to acknowledge that cancer diagnosis affects death through other mechanisms independent of $P$

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```

---
class: middle, center

```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->] (dp) -- (pstar);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

...all the pieces together...

@@

We conduct an analysis of the cancer-ADRD association that is structured to the best of the available data’s abilities to inform the possible effect of Pin1 on all cause-dementia

- Participants between 60 and 70 years old at study entry
- Without history of cancer diagnosis and dementia
- Data on cancer diagnosis and dementia diagnosis is measured through follow-up and linkage with national registries.
- Date and cause of death was collected on a weekly basis via municipal population registries
- Data on covariates is measured repeatedly over visits and through linkage to EHR

@@

We considered 3 scenarios to define Pin1 over-expression:

@@

- _"Cancer ever vs. never"_ as if Pin1 measurement and cancer diagnosis happened at the same time if looked retrospectively.

@@

- Cancer as a _"time-varying"_ exposure independent of death.

- _"Time to cancer diagnosis"_ as the proxy for Pin1 over-expression.

We used inverse probability treatment weighting (IPTW) to address confounding.

- To compare the risk of dementia at 20 years of follow-up, in a setting were death could have been prevented, we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs. no incident of cancer with time indexed in years.

- The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring by death weights (IPC).

- For an individual still alive in year _t_, the time _t_ IPC weight is the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia. For an individual who has died by time t, the year t IPC weight is zero.

- We additionally calculated the direct effect considering death as an unconditional independent censoring event (as if there were no arrows from {C} to $D_{t}$ and $Y_{t+1}$) for an illustrative purpose.

@@

- Among participants free of cancer diagnosis, 14% had dementia and 23% died.
]
---
### Results

*Weight assessment for cancer diagnosis*

.center[
```{r}
love.plot(w.out1) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#011A5E", "#e4a803"))

```
]

*Weight assessment for death including baseline covariates*

.center[
```{r}
love.plot(w.out_death) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#011A5E", "#e4a803"))
```
]
---
### Cancer as: never vs. ever

@@

---
### Cancer as time-varying exposure

.pull-left[
```{r}
plot_2b
```

.pull-right[
```{r}
plot_2c
```

@@

.center[
```{r, echo = FALSE}
tb <- table_results %>%
  select(-c(3, 4)) %>%
  rename(RD = rd, RR = rr, `HR (95%CI)` = hr)

@@

### Extreme scenarios of independency

@@

### Causes of death

```{r, echo = FALSE, fig.align='center', fig.width = 14}
causes_death + 
      theme_xaringan(text_font_size = 20,
                       title_font_size = 20) +
     theme(legend.position = "right")

### Thoughts so far...

- There big gap between what we measure and analyze as exposures, and what we truly want to study.

@@

@@

- Descriptive information, such as % of people who died in each arm should be always presented.

@@

### Questions

- How can I make my points clear for a wide audience?

- What to add, what to remove?

>>>>

---
title: "Toward a clearer understanding of the inverse association between cancer and dementia"
subtitle: ""  
author: 
  - "L. Paloma Rojas-Saunero"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["r", "css"]
      countIncrementalSlides: FALSE
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

@@

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#011A5E",
  header_font_google = google_font("Fira Sans", "500"),
  table_row_even_background_color = "#FFFFFF",
  table_border_color = "#011A5E",
  text_font_google   = google_font("Lato", "400", "400i"),
  text_font_size = "30px"
)
```

background-image: url(./figs/ser.jpg), url(./figs/erasmus.png)
background-position: 5% 95%, 95% 5%
background-size: 25%, 15%
class: center, middle

## Towards a clearer understanding of the inverse association between cancer and dementia

@@

---
## Introduction

- Observational studies have found an inverse association between cancer and the risk of Alzheimer´s disease or related dementias (ADRD).

@@

- To understand the potential sources of bias we first need to define the causal question.

- Refining the underlying question behind cancer diagnosis as the exposure unveils some of these concerns.

@@

To conceptualize and progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.

---
### Pin1-targeting drug

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$D$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (p) -- (y);
\end{tikzpicture}
```

What is the effect of a Pin1 targeting drug $(D)$ in the risk of ADRD over time *t* $(Y_{t})$  compared to standard treatments?

@@

???
In hypothetical setting, if Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a randomized trial in which we randomize eligible participants in late midlife to receive this drug or not, and closely monitor ADRD over follow-up.

---
### Pin1 expression as biomarker

@@

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (l) -- (p);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $L$ = shared factors between $P$ and $Y_t$

@@

???
We might be closer to developing biomarker for Pin1 expression in the future. So we can imagine a situation in which Pin1 exposure is measured in a population based cohort study, using stored blood samples.
Since this would be a observational setting, Pin1 expression and ADRD at time may share causes such as smoking, and that assessing the causal relationship requires adjusting for these confounders.

---
### Cancer diagnosis as the proxy of Pin1 biomarker

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->, violet] (p) -- (pstar);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);

\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $P^*$ = Cancer diagnosis; $L$ = shared factors between $P$ and $Y_t$ = ADRD diagnosis
]

???
At this moment this biomarker is not available, but we know that Pin1 over-expression stimulates tumor growth and for this reason we use cancer diagnosis as the proxy for Pin1.

---
### Zoom in the P and P* association

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\end{tikzpicture}
```

<br> <br>

$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$ ;

@@

---
### Selection bias in the P and P* association

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) -- (d);
\draw[->] (c) -- (pstar);
\draw[->] (c) -- (d);
\end{tikzpicture}
```

<br> <br>

$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$; $C_1$, $C_2$ shared factors such as smoking

@@

???
In some way we need to assume that those who died would have the same risk of having a cancer diagnosis had they not died 
---
### Direct effect vs. total effect, potential bias

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->, very thick] (d) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```
]

@@

---
### All pieces together

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->] (dp) -- (pstar);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

At best we can test the sharp null hypothesis
]

@@

- Participants between 60 and 70 years old at study entry.

- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and electronical clinical records.

Three definitions of the proxy for Pin1 over-expression:

- _"Cancer ever vs. never"_

- Cancer as a _"time-varying"_

- _"Time to cancer diagnosis"_

@@

- Inverse probability treatment weights to address confounding

  + *Age, sex, education, smoking status*

@@

- Inverse probability censoring weights for death to relax the independent censoring assumption

  + *Age, sex, Apoe4, education and time-varying covariates: smoking, systolic blood pressure, BMI and incident cancer, heart disease, stroke and diabetes*

@@

- We estimate the risk of dementia using a weighted Kaplan-Meier estimator.

@@

- Among participants free of cancer diagnosis, 14% had dementia and 23% died.
]
---
### Cancer as: never vs. ever

@@

@@

.center[
```{r, echo = FALSE}
tb <- table_results %>%
  filter(model != "Unadjusted") %>% 
  select(-c(3, 4)) %>%
  rename(RD = rd, RR = rr, `HR (95%CI)` = hr)

@@

### Extreme scenarios of independency between death and dementia

@@

---
### Conclusions

@@

- Cancer as the proxy for a molecular mechanism is tricky...
--

@@

@@
